---
title: "Forecasting 3"
author: "Ryan Ochoa"
date: "3/13/2022"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
setwd("C:\\Users\\rocho\\OneDrive\\Documents\\R\\Forecasting\\Homework\\Problem Set 3")
library(fma)
library(forecast)
library(ggplot2)
library(fpp2)
library(tseries)
```



## (1) Explain the differences among these figures. Do they all indicate the data are white noise? Why are the critical values at different distances from the mean of zero? 


The data provide by each of the three graphs shows that all lags are within the confidence interval, thus indicating that each data is white noise.


## (2a)	Use the following R code to generate data from an AR(1) model with ϕ1=0.6 and σ2=1. The process starts with y0=0.  Produce a time plot for the series.


```{r, echo=FALSE}
y <- ts(numeric(100))
e <- rnorm(100)
for(i in 2:100)
  y[i] <- 0.6*y[i-1] + e[i]

ts.plot(y)
```





## (2b)	How does the plot change as you change ϕ1 (0.3, 0.6 and 0.9)?




```{r, echo=FALSE}
#ϕ1=0.3
y1 <- ts(numeric(100))
e <- rnorm(100)
for(i in 2:100)
  y1[i] <- 0.3*y1[i-1] + e[i]

ts.plot(y1)
```



```{r, echo=FALSE}
#ϕ1=0.9
y2 <- ts(numeric(100))
e <- rnorm(100)
for(i in 2:100)
  y2[i] <- 0.9*y2[i-1] + e[i]

ts.plot(y2)
```





## (3a)	Plot the data against time and obtain ACF, PACF and Q-statistics.

**-	Does the series appear to be stationary**?


```{r, echo=FALSE}
y2 <- read.csv("Y2_new.csv", header=TRUE)
yy <- ts(y2)
plot(yy)
```


From the graph above, we do not see any trend or seasonality. Visually, we can conclude that the data is *stationary*.


```{r, echo=FALSE}
ggtsdisplay(yy,main="")
```


**-	Does the series appear to be a white noise process?  Why or why not?**

```{r, echo=FALSE}
#Q*test
Box.test(yy, lag=24, fitdf=0, type="Ljung")
```

**p-value < 0.05**, therefore the data is not white noise.



## (3b)	Estimate different ARMA models using various values of p and q from 0 to 5, and identify the best ARMA(p,q) model (use AICc information criteria).Then, provide the estimation result of the best ARMA model.

Using the *auto.arima()* function, R produces a (1,0,1) model.

```{r, echo=FALSE}
autofit <- auto.arima(yy,seasonal=FALSE)  
summary(autofit)

```


Using all varying values for p and q form 0 to 5, different ARMA models are estimated to identify the best model using a AICc criteria:

*table*

```{r, include=FALSE}
fit1 <- Arima(yy, order=c(0,0,1), include.mean=FALSE)
fit2 <- Arima(yy, order=c(0,0,2), include.mean=FALSE)
fit3 <- Arima(yy, order=c(0,0,3), include.mean=FALSE)
fit4 <- Arima(yy, order=c(0,0,4), include.mean=FALSE)
fit5 <- Arima(yy, order=c(0,0,5), include.mean=FALSE)
fit6 <- Arima(yy, order=c(1,0,1), include.mean=FALSE)
fit7 <- Arima(yy, order=c(1,0,2), include.mean=FALSE)
fit8 <- Arima(yy, order=c(1,0,3), include.mean=FALSE)
fit9 <- Arima(yy, order=c(1,0,4), include.mean=FALSE)
fit10 <- Arima(yy, order=c(1,0,5), include.mean=FALSE)
fit11 <- Arima(yy, order=c(2,0,1), include.mean=FALSE)
fit12 <- Arima(yy, order=c(2,0,2), include.mean=FALSE)
fit13 <- Arima(yy, order=c(2,0,3), include.mean=FALSE)
fit14 <- Arima(yy, order=c(2,0,4), include.mean=FALSE)
fit15 <- Arima(yy, order=c(2,0,5), include.mean=FALSE)
fit16 <- Arima(yy, order=c(3,0,1), include.mean=FALSE)
fit17 <- Arima(yy, order=c(3,0,2), include.mean=FALSE)
fit18 <- Arima(yy, order=c(3,0,3), include.mean=FALSE)
fit19 <- Arima(yy, order=c(3,0,4), include.mean=FALSE)
fit20 <- Arima(yy, order=c(3,0,5), include.mean=FALSE)
fit21 <- Arima(yy, order=c(4,0,1), include.mean=FALSE)
fit22 <- Arima(yy, order=c(4,0,2), include.mean=FALSE)
fit23 <- Arima(yy, order=c(4,0,3), include.mean=FALSE)
fit24 <- Arima(yy, order=c(4,0,4), include.mean=FALSE)
fit25 <- Arima(yy, order=c(4,0,5), include.mean=FALSE)
fit26 <- Arima(yy, order=c(5,0,1), include.mean=FALSE)
fit27 <- Arima(yy, order=c(5,0,2), include.mean=FALSE)
fit28 <- Arima(yy, order=c(5,0,3), include.mean=FALSE)
fit29 <- Arima(yy, order=c(5,0,4), include.mean=FALSE)
fit30 <- Arima(yy, order=c(5,0,5), include.mean=FALSE)
fit31 <- Arima(yy, order=c(0,0,0), include.mean=FALSE)
fit32 <- Arima(yy, order=c(1,0,0), include.mean=FALSE)
fit33 <- Arima(yy, order=c(2,0,0), include.mean=FALSE)
fit34 <- Arima(yy, order=c(3,0,0), include.mean=FALSE)
fit35 <- Arima(yy, order=c(4,0,0), include.mean=FALSE)
fit36 <- Arima(yy, order=c(5,0,0), include.mean=FALSE)
summary(fit1)
summary(fit2)
summary(fit3)
summary(fit4)
summary(fit5)
summary(fit6) #AICc=2835.81
summary(fit7)
summary(fit8)
summary(fit9)
summary(fit10)
summary(fit11)
summary(fit12)
summary(fit13)
summary(fit14)
summary(fit15)
summary(fit16)
summary(fit17)
summary(fit18)
summary(fit19)
summary(fit20)
summary(fit21)
summary(fit22)
summary(fit23)
summary(fit24)
summary(fit25)
summary(fit26)
summary(fit27)
summary(fit28)
summary(fit29) #AICc=2833.87
summary(fit30) #AICc=2831.32
summary(fit31)
summary(fit32)
summary(fit33)
summary(fit34)
summary(fit35)
summary(fit36)
```






## (3c)	Provide diagnostic check-ups for the residuals of the model you estimated by examining the ACF, PACF and Q-statistics.  Does the residual series appear to be a white noise process?


The following tests will provide diagnostic check-ups for the residuals of the model for the three possible models found in the previous section: (1,0,1), (5,0,4), and (5,0,5).

```{r, echo=FALSE}
#fit6; (1,0,1)
ggAcf(residuals(fit6))
checkresiduals(fit6)

#fit29; (5,0,4)
ggAcf(residuals(fit29))
checkresiduals(fit29)

#fit30; (5,0,5)
ggAcf(residuals(fit30))
checkresiduals(fit30)
```


For fit6; (1,0,1): p-value=0.1599 > 0.05, FTR the null of white noise.  **Residuals are white noise**.

For fit29; (5,0,4): p-value=0.03561 < 0.05, Reject the null of white noise. Residuals are not white noise.

For fit30: p-value=0.02342 < 0.05, Reject the null of white noise. Residuals are not white noise.




## (3d)	Obtain the out-of-sample forecasts of the next 20 periods using the estimated ARMA model. Plot the data and the forecasts.

20-period forecast using the (1,0,1) model:

```{r, echo=FALSE}
forecast(fit6, h=20)
plot(forecast(fit6, h=20))
```


## (4a) Plot the data against time and obtain ACF, PACF and Q-statistics.  Does the series appear to be stationary?  (Note that the series would be non-stationary when the ACFs decay slowly.)

```{r, echo=FALSE}
us <- read.csv("us.csv", header=TRUE)
gdp <- us[,"GDP85"]
gdp.ts <- ts(gdp)
ts.plot(gdp.ts)
```

From the above graph, we can visually conclude that the data is *non-stationary*.

```{r, echo=FALSE}
#confirmation using test for stationarity
ggtsdisplay(gdp.ts,main="")
adf.test(gdp.ts, alternative = "stationary") #Reject null of unit root, thus data is non-stationary
```

p-val = 0.1706 > 0.05; FTR the null hypothesis of "non-stationarity". 

From the ACF graph above, we can see a slow and sluggish geometric decay, providing more evidence that the data is non-stationary.  


p-val = 0.1706 > 0.05; FTR the null hypothesis of "non-stationarity". 

Furthermore, by using the Augmented Dickey-Fuller Test, we can confirm that the data is indeed *non-stationary*  


## (4b)	You must have concluded that the series appears non-stationary.  To make it stationary, first difference the logged series and call it as GDP_FD: GDP_FD = dlog(GDP85)t = log(GDP85)t - log(GDP85)t-1 



```{r, echo=FALSE}
ndiffs(gdp.ts, alpha=0.05, test=c("kpss","adf", "pp"), max.d=2)
gdp.ts.diff <- diff(log(gdp.ts), differences=1)
```



**-	Plot this detrended data against time and obtain ACF, PACF and Q-statistics.**  


```{r}
ggtsdisplay(gdp.ts.diff,main="")
```


**-	Does the series appear to be stationary?**  


```{r}
adf.test(gdp.ts.diff, alternative = "stationary")
```

From the Augmented Dickey-Fuller Test, the p-value=0.01 < 0.05, thus we 

**-	Does the series appear to be a white noise process?**


```{r}
#white noise?
Box.test(gdp.ts.diff, lag=24, fitdf=0, type="Ljung") 
#p-val=0.02704, not white noise
```

pivalue=0.02704<0.05; thus reject null of white noise, thus the data is not a white noise process


## (4c)	Provide the estimation result of the best ARMA model. Then, provide diagnostic check-ups for the residuals of the model you estimated by examining the ACF, PACF and Q-statistics.  Does the residual series appear to be a white noise process?

The residual p-val > 0.05, thus the residuals are white noise.


```{r}
fitty <- auto.arima(gdp.ts.diff,seasonal=FALSE)  
summary(fitty)
ggAcf(residuals(fitty))
checkresiduals(fitty)
```


## (4d)	Obtain the out-of-sample forecasts of the next 10 quarters using the estimated ARMA model. Plot the data and forecasts.

10-period forecast using the (2,0,0) model:

```{r, echo=TRUE}
forecast(fitty)
plot(forecast(fitty))
```



